### 2015-01-24 doc/thesis/system.tex
Updating typing rules to use the projections environment

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of the tables that have no keys that do not
inhabit one of the table's key types, and that have no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n_{e} \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m} \; | \;
e^{x}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

We present the typing rules as a deduction system for the typing
relations $\env_{1}, \penv \vdash s, \env_{2}$ and
$\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
The second relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify the environment.
We need two environments because $\env$ maps variables to types
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[\pi \mapsto r]$ for meaning that we extendend the environment
$\penv$ with the projection $\pi$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consitent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

@@

We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
to the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinment of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

The rule \textsc{T-EXPLIST2} is close to the rule \textsc{T-LHSLIST}.
The difference is that it uses $\Nil{*}$ to replace missing values.
We define this rule as follows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env \vdash e_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

The typing rules that type check expression lists include $\Nil{*}$
in the end of the tuple type only when it does not end in a variadic type.
For instance, the rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4}
handle the case that an expression list is just an expression that
return multiple values, but only \textsc{T-EXPLIST3} needs to add
$\Nil{*}$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
      {\env \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
      {\env \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
\end{array}
\end{array}
\]

As an example, the rule \textsc{T-EXPLIST3} allows our type system
to type check the following example:
\[
x, y, z = f()_{m}
\]
assuming that $x$, $y$, $z$, and $f$ are in the environment with
types $\Integer$, $\String$, $\String \cup \Nil$, and
$\Void \rightarrow \Integer \times \String \times \Void$, respectively.
It type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

@@

Multiple assignments also appear in function applications, and the
rule \textsc{T-APPLY1} uses the same behavior from multiple assignments
to drop extra parameters and replace missing parameters:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{3} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

Assuming that $x$ and $f$ are local variables that are in the environment
with types $\Integer \cup \Nil$ and
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
respectively.
The program
\[
x = f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1} because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the program
\[
x = f(``foo",1,2,3)_{m}
\]
also type checks through the rule \textsc{T-APPLY1} because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

However, we also expect from a type system that it catches arity mismatch.
For this reason, our type system also has different typing rules for
expression lists that appear in function applications.
These rules do not include $\Nil{*}$ when the tuple type does not end
in a variadic type.
The rule \textsc{T-APPLY1-STRICT} use these rules to type check
expression lists, and it also uses the predicate $adjust$ to replace
any missing value with $\Nil$:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1} \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2} \vdash el:r_{3}, \env_{3} \;\;\;
       r_{4} = adjust(r_{3}, r_{1}) \;\;\;
       r_{4} \lesssim r_{1}}
      {\env_{1} \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

We can use pattern matching to the define the recursive predicate
\emph{adjust} as follows:
\begin{align*}
adjust(\Void, t{*}) & = \Nil \times \Void\\
adjust(\Void, t \times s) & = \Nil \times adjust(\Void,s)\\
adjust(t_{1} \times s_{1}, t_{2} \times s_{2}) & = t_{1} \times adjust(s1,s2)\\
adjust(s_{1}, s_{2}) & = s_{1}
\end{align*}

Now, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the program
\[
x = f(``foo")_{m}
\]
type checks through the rule \textsc{T-APPLY1-STRICT} because
\[
``foo" \times \Nil \times \Nil \times \Void \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the program
\[
x = f(``foo",1,2,3)_{m}
\]
does not type check through the rule \textsc{T-APPLY1-STRICT} because
\[
``foo" \times 1 \times 2 \times 3 \times \Void \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

The same happens with the rules that type check the \textbf{return}
statement.
The rule \textsc{T-RETURN} uses $\Nil{*}$ to replace eventual missing
values, while the rule \textsc{T-RETURN-STRICT} uses the predicate
\emph{adjust} to replace values that are certainly missing:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-RETURN}\\
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}(\ret) = r_{2} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-RETURN-STRICT}\\
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}(\ret) = r_{2} \;\;\;
       r_{3} = adjust(r_{1}, r_{2}) \;\;\;
       r_{3} \lesssim r_{2}}
      {\env_{1} \vdash \mathbf{return} \; el:\env_{2}}
\end{array}
\end{array}
\]

@@

The simplest form of the table constructor is the empty table $\{\}$.
Its rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of table fields $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
We chose to not include the latter form because its type checking rule is similar
to the type checking rule of the first form, and just controls the type of
the omitted indices.
For the same reason we did not include table constructors with
multiple expressions, which can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can type check a table type that has a record part and
also an array part.

The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i} \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env_{1} \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

@@

The type checking of the table fields would be straightforward
if our table types allowed any first-level type in the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-FIELD4} is a fallback for the previous rules:
\[
\begin{array}{c}
\mylabel{T-FIELD4}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3}}
      {\env_{1} \vdash [e_{1}] = e_{2}: \Value{:}close(t_{2}), \env_{3}}
\end{array}
\]

>>>>

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of tables with no keys that do not
inhabit one of the table's key types, and with no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n_{e} \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

@@

We present the typing rules as a deduction system for the typing
relations $\env_{1}, \penv \vdash s, \env_{2}$ and
$\env_{1}, \penv \vdash e : t, \env_{2}$.
We use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
typing the statement $s$ produces the new type environment $\env_{2}$.
The second relation means that given the type environment $\env_{1}$
and the projection environment $\penv$,
the expression $e$ has type $t$ and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though expressions and statements can modify the environment.
We need two environments because $\env$ maps variables to types,
while $\penv$ maps projection variables to unions of tuple types.
We use $\env_{1}[n \mapsto t]$ for meaning that we extended the environment
$\env_{1}$ with the variable $n$ that maps to type $t$.
We use $\penv[x \mapsto r]$ for meaning that we extendend the environment
$\penv$ with the projection variable $x$ that maps to the union of tuple types $r$.

Lua has multiple assignments, and the rule \textsc{T-ASSIGNMENT} uses
tuple types along with consitent-subtyping to type check this Lua feature:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \vec{l} = el,\env_{3}}
\end{array}
\end{array}
\]

@@

We define the rule \textsc{T-LHSLIST} as follows:
\[
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{l}\;|}
      {\env, \penv \vdash \vec{l}:t_{1} \times ... \times t_{n} \times \Value{*}, \env_{m}}
\end{array}
\]

First, this rule uses the same environment $\env$ to type check each
left-hand side value $l_{k}$, as they can perform different changes
in the environment.
Then, it takes each type $t_{k}$ and builds the tuple type for
the assignment rule.
It places the type $\Value{*}$ in the end to discard extra values.
The rule succeeds if the predicate \emph{merge} can
produce a new environment, which includes all the modifications
of each environment $\env_{k}$.
Intuitively, the predicate \emph{merge} fails when an environment
tries to change the type of a variable towards a type that is not
a subtype of its previous type.
We will show in this section that the refinment of table types can
change the environment, and that is the reason why we need to check
whether the environment is consistent.

The rule \textsc{T-EXPLIST2} is close to the rule \textsc{T-LHSLIST}.
The difference is that it uses $\Nil{*}$ to replace missing values.
We define this rule as follows:
\[
\begin{array}{c}
\mylabel{T-EXPLIST2}\\
\dfrac{\env, \penv \vdash e_{k}:t_{k}, \env_{k} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\vec{e}\;|}
      {\env, \penv \vdash \vec{e}:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{m}}
\end{array}
\]

The typing rules that type check expression lists include $\Nil{*}$
in the end of the tuple type only when it does not end in a variadic type.
For instance, the rules \textsc{T-EXPLIST3} and \textsc{T-EXPLIST4}
handle the case that an expression list is just an expression that
return multiple values, but only \textsc{T-EXPLIST3} needs to add
$\Nil{*}$:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-EXPLIST3}\\
\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Void, \env_{1}}
      {\env, \penv \vdash me:t_{1} \times ... \times t_{n} \times \Nil{*}, \env_{1}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPLIST4}\\
\dfrac{\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
      {\env, \penv \vdash me:t_{1} \times ... \times t_{n}{*}, \env_{1}}
\end{array}
\end{array}
\]

As an example, the rule \textsc{T-EXPLIST3} allows our type system
to type check the following example:
\[
x, y, z = f()_{m}
\]

In this example, we are assuming that $x$, $y$, $z$, and $f$ are in
the environment with types $\Integer$, $\String$, $\String \cup \Nil$, and
$\Void \rightarrow \Integer \times \String \times \Void$, respectively.
Thus, the example type checks because
\[
\Integer \times \String \times \Nil{*} \lesssim \Integer \times \String \times \String \cup \Nil \times \Value{*}
\]

@@

Multiple assignments also appear in function applications.
The rule \textsc{T-APPLY1} handles the case where function applications
are expressions that return multiple values:
\[
\begin{array}{c}
\mylabel{T-APPLY1}\\
\dfrac{\env_{1}, \penv \vdash e:r_{1} \rightarrow r_{2}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash el:r_{3}, \env_{3} \;\;\;
       r_{3} \lesssim r_{1}}
      {\env_{1}, \penv \vdash e(el)_{m}:r_{2}, \env_{3}}
\end{array}
\]

We also use the rule \textsc{T-APPLY1} as the base case for the rules
that handle the cases where function applications are either statements
or expressions that return only one value:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-STMAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e(el)_{s},\env_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-EXPAPPLY1}\\
\dfrac{\env_{1}, \penv \vdash e(el)_{m}:r, \env_{2}}
      {\env_{1}, \penv \vdash e(el)_{e}:first(r), \env_{2}}
\end{array}
\end{array}
\]

The rule \textsc{T-STMAPPLY1} discards the returned values,
while the rule \textsc{T-EXPAPPLY1} uses the predicate \emph{first} to
ensure that only one value is returned.
We can use pattern matching to the define the recursive predicate
\emph{first} as follows:
\begin{align*}
first(\Void) & = \Nil\\
first(t{*}) & = t \cup \Nil\\
first(t \times s) & = t\\
first(s_{1} \sqcup s_{2}) & = first(s_{1}) \cup first(s_{2})
\end{align*}

Assuming that $f$ is a local function in the environment, and that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*} \rightarrow \Integer{*}$,
the function call
\[
f(``foo")_{s}
\]
type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]
and the function call
\[
x = f(``foo",1,2,3)_{m}
\]
also type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Value{*}
\]

Our type system also catches arity mismatch.
For instance, assuming that $f$ has type
$\String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void \rightarrow \Integer{*}$,
the function call
\[
f(``foo")_{s}
\]
type checks through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
``foo" \times \Nil{*} \lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]
but the function call
\[
x = f(``foo",1,2,3)_{m}
\]
does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
``foo" \times 1 \times 2 \times 3 \times \Nil{*} \not\lesssim \String \times \Integer \cup \Nil \times \Integer \cup \Nil \times \Void
\]

When our type system type checks an expression list,
it always include $\Nil{*}$ in the end of the type of this expression list,
if the type does not end in a variadic type.
Our type system has this behavior because $\Nil{*}$ along with subtyping
and consistent-subtyping use $\Nil$ to replace missing values.
This behavior preserves the semantics of Lua on replacing missing values,
as it is necessary when we ommit optional parameters in a function call,
like the previous example showed.

Using $\Nil{*}$ in the end of the type of expression lists also allows
our type system to catch arity mismatch in function calls where
the type of the function does not include optional parameters.
For instance, assuming that $f$ has type
$\Integer \times \Integer \times \Void \rightarrow \Integer \times \Void$,
the function call
\[
f(1)_{s}
\]
does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
1 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]
and the function call
\[
x = f(1,2,3)_{m}
\]
also does not type check through the rules \textsc{T-STMAPPLY1} and \textsc{T-APPLY1}, because
\[
1 \times 2 \times 3 \times \Nil{*} \not\lesssim \Integer \times \Integer \times \Void
\]

The same behavior happens with the rule that type checks the \textbf{return}
statement:
\[
\begin{array}{c}
\mylabel{T-RETURN}\\
\dfrac{\env_{1}, \penv \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2}(\ret) = r_{2} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1}, \penv \vdash \mathbf{return} \; el,\env_{2}}
\end{array}
\]

@@

The simplest form of the table constructor is the empty table $\{\}$.
Its rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env, \penv \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of table fields $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions, because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
We chose to not include the latter form because its type checking rule is similar
to the type checking rule of the first form, and it just controls the type of
the omitted indices.
For the same reason, we did not include table constructors with
multiple expressions, which can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can type check a table type that has a record part and
also an array part.

The rule \textsc{T-CONSTRUCTOR2} type checks a table type of the first form:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:k_{i}{:}v_{i}, \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       t = wf(\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique})}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:t, \env_{n+1}}
\end{array}
\]

@@

The type checking of the table fields would be straightforward
if our table types allowed any first-level type in the type of the keys.
The rules \textsc{T-FIELD1}, \textsc{T-FIELD2}, and \textsc{T-FIELD3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-FIELD1}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD2}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-FIELD3}\\
\dfrac{\env_{1}, \penv \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1}, \penv \vdash [e_{1}] = e_{2}: t_{1}{:}close(t_{2}), \env_{3}}
\end{array}
\end{array}
\]

These rules use the predicate \emph{close} in the type of the value
because our type system do not allow \emph{open} and \emph{unique}
table types to appear in the type of the values.
The predicate \emph{close} promotes all table types to \emph{closed}.
We made this restriction because our type system does not keep track
of aliases to table fields.
This means that allowing \emph{open} and \emph{unique} table types
to appear in the type of the values would allow the creation of
unsafe aliases.
